A Multi-Fidelity Framework for Wildland Fire Behavior Simulations over Complex Terrain

### 2021-01-07 Vanella_Article_Atmosphere_2020.tex
Update on Overleaf.

%=================================================================
% Full title of the paper (Capitalized)
\Title{Application of a Cut-Cell Immersed Boundary Method for Wildfire Simulation over Complex Terrain}

@@

There are various ways of simulating wildland fire spread depending on the resolution of the underlying numerical grid. 
\begin{enumerate}
\item {\bf Particle Model:} The vegetation is represented by a collection of Lagrangian particles that are heated via convection and radiation. This model is appropriate for grid resolution of the order of 1~m or less.
\item {\bf Boundary Fuel Model:} Ground vegetation is modeled like a porous solid with a thickness equal to the height of the vegetation. This model is appropriate for grid resolution of the order of 1~m to 10~m.
\item {\bf Level Set Method:} The fire front propagates using purely empirical rules. This model is appropriate for grid resolution of the order of roughly 10~m and greater. 
\end{enumerate}
The Particle Method and Boundary Fuel Model require thermo-physical properties of the vegetative fuels and the fire spread rate is {\em predicted} by the model. The Level Set Method relies on a set of predetermined spread rates for different types of vegetation and wind speeds. The pyrolysis model, developed specifically for vegetation~\cite{Porterie:2006,Morvan:CF2004,Houssami:2016}, consists of three reactions:

@@

In many simulations of wildland fire, the ground vegetation layer is too shallow to be resolved explicitly, as is done when using Lagrangian particles to represent the vegetation. In such cases, the ground vegetation can be modeled as a porous boundary consisting of a layer of dry vegetation, moisture, and air, underneath which is hard ground~\cite{Perez-Ramirez:FT2017}. The drag exerted by the vegetation is modeled using a special velocity boundary condition, and convective heat transfer is modeled via a source term in the one-dimensional heat conduction equation that is solved through the layer of vegetation and solid ground. Thermal radiation penetrates the vegetation layer via a 1-D radiative transport equation that is used for semi-transparent solids.

@@

For simulations of wildland fires spanning large areas that cannot be gridded finely enough to predict fire spread using a physics-based model, fire spread via level sets is an option~\cite{Bova:IJWF2015}. This approach makes use of the Rothermel-Albini~\cite{Rothermel:1972,Albini:1976} surface fire spread rate formulae and the assumption that a surface fire spreading from a point under certain wind, slope and vegetation conditions does so with an ellipse-shaped fire front with, for a given wind speed, a fixed length-to-breadth ratio~\cite{Richards:1990}. The location of a fire front at time $t$ is defined by the set of points $x,y$ where the level set scalar function $\phi(x,y,t)=0$. The advection equation for the level set function is
\begin{equation}
   \frac{D \phi(x,y,t)}{D t} =  \frac{\partial \phi}{\partial t} + \mathbf{R}_{uv} \cdot \nabla (\phi) = 0 \label{eqn:lset}
\end{equation}
where $\mathbf{R}_{uv}=(R_u,R_v)$ is the spread rate vector. The magnitude of the vector is
\begin{equation}
  \|\mathbf{R}_{uv}\|=R_0 \left(1 + \sqrt{(\boldsymbol{\phi}_W+\boldsymbol{\phi}_S) \cdot (\boldsymbol{\phi}_W+\boldsymbol{\phi}_S) } \right)
\end{equation}
where $R_0$ is the zero velocity, zero slope rate of spread that depends on the fuel properties. The vectors $\boldsymbol{\phi}_W,\boldsymbol{\phi}_S \ge 0$ are computed using the local wind and slope through empirical rules~\cite{Wilson:1980}. See Appendix 1 of Ref~\cite{Bova:IJWF2015} for details on these calculations. In FDS, the level set spread model can be applied in various ways:
\begin{enumerate}
\item Only the level set simulation is performed, with a constant and uniform specified wind and slope. The wind is not affected by the terrain, and there is no fire.
\item The wind field is established over the terrain, but it is ``frozen'' when the fire ignites.
\item The wind field follows the terrain, but there is no actual fire in the simulation, just front-tracking.  The level set evolves continuously in time with the flow field.
\item The wind and fire are fully-coupled. When the fire front arrives at a given surface cell, it burns for a finite duration and with a heat release per unit area provided as part of the fuel model.
\end{enumerate}
When fully coupled to the CFD model, the level set function acts as an igniter as it reaches a new surface cell. The cell burns for a pre-determined amount of time. For cases where the depth of the fireline is less than the size of a grid cell, the burning time is extended while the burning rate is decreased so as to maintain conservation of fuel mass.

@@

For outdoor flows FDS prescribes the wind field through special ``open'' boundary conditions on the exterior of the computational domain.  The boundary conditions consist of three components:
\begin{enumerate}
\item[(i)] the specified upstream wind field (vertical profile of streamwise velocity components) based on prescribed Monin-Obukhov parameters imposed on fluid elements \emph{entering} the domain,
\item[(ii)] optional specification of upstream turbulence based on Jarrin's synthetic eddy method \cite{Jarrin:2008} (which is possible with the code, but not utilized in the test cases within this paper), and
\item[(iii)] nonuniform and nonstationary Dirichlet pressure boundary values for the Poisson equation.
\end{enumerate}

@@

\subsection{Turbulent Inflow Conditions}

It is possible in FDS to superimpose synthetic turbulence on a specified inlet profile utilizing Jarrin's synthetic eddy method (SEM) \cite{Jarrin:2008}.  An example verification case is presented in Fig.~\ref{fig:semprofs}.  The left image of the figure shows contours of the normal component of velocity at the inlet.  The plot on the right shows the mean and rms profiles compared to the specified inlet parameters.  The domain is 20 m in the streamwise dimension by 12 m in the spanwise by 13 m in height.  In The turbulence intensity is set to 10 \% and the turbulence integral length scale is set to 3 m.  The specified MO parameters are $L=-667\;\mathrm{m}$, $z_0=0.022\;\mathrm{m}$, and $u_r=10\;\mathrm{m/s}$ and $T_r=20\;^\circ\mathrm{C}$ at reference height of 10 m.  The simulation runs for 60 seconds with statistics collected over the last 30 seconds.  The lateral, top, and downstream boundaries are set to ``open'' pressure boundaries as discussed in the next section.

\begin{figure}[ht]
\centering
\includegraphics[width=0.5\textwidth,valign=c]{figures/sem_open_wind_contours.pdf}
\includegraphics[width=0.48\textwidth,valign=c]{figures/sem_open_wind_u_prof.pdf}
\caption{Example of inflow synthetic turbulence superimposed on a Monin-Obukhov boundary layer profile. Contours of the normal component of velocity at the domain inlet taken at a time midway through the simulation (left); mean and rms profiles along the inlet compared with specified profiles (right).}
\label{fig:semprofs}
\end{figure}

Other approaches to imposing boundary layer turbulence are possible (e.g., \cite{Munoz-Esparza:2014}).  The cases presented later in this paper do not utilize SEM since detailed data on the turbulence statistics were not available.

@@

\begin{table}[ht]
\begin{center}
\caption[Measured properties for the CSIRO Grassland Fire cases]{Measured properties for the CSIRO Grassland Fire cases~\cite{Cheney:IJWF1993}.}
\label{Properties_Grasses}
\begin{tabular}{|l|c|c|c|}
\hline
Property                        & Units        & Case C064 & Case F19    \\ \hline \hline
Wind Speed                      & m/s          & 4.6       & 4.8         \\ \hline
Ambient Temperature             & $^\circ$C    & 32.       & 34          \\ \hline
Surface Area to Volume Ratio    & m$^{-1}$     & 9770      & 12240       \\ \hline
Grass Height                    & m            & 0.21      & 0.51        \\ \hline
Bulk Mass per Unit Area         & kg/m$^2$     & 0.283     & 0.313       \\ \hline
Moisture Fraction               & \%           & 6.3       & 5.8         \\ \hline
Measured RoS                    & m/s          & 1.1       & 1.5         \\ \hline
Calc'd RoS, Particle Method (0.25 m, 0.5 m, 1.0 m resolution) & m/s           & 1.4, 1.3, 1.4           & 1.1, 1.2, 1.2          \\ \hline
Calc'd RoS, Boundary Fuel Method (0.25 m, 0.5 m, 1.0 m resolution) & m/s      & 1.5, 1.4, 1.7           & 1.3, 1.3, 1.3           \\ \hline
Calc'd RoS, Level Set Method (5 m, 10 m, 20 m resolution) & m/s          & 1.1, 1.1, 1.2          & 0.4, 0.5, 0.8          \\ \hline
\end{tabular}
\end{center}
\end{table}

@@

The simulations have been conducted with the three different fire spread models, and each case has been run with three levels of spatial resolution. The simulations using the Particle and Boundary Fuel Models have been run with cubic grid cells that are 0.25~m, 0.5~m, and 1~m on a side. The simulations using the Level Set Model are run with cells of 5~m, 10~m, and 20~m. The predicted rates of spread for the two cases are listed in Table~\ref{Properties_Grasses} and shown graphically in Fig.~\ref{CSIRO}. For these simulations, the rate of spread predicted by the Particle and Boundary Fuel Models are comparable to the measured rate of spread while the level set method under-predicts the RoS. The under-prediction by the level set simulations is a consequence of the choice of fuel model, and one should not draw any conclusions as to the performance of level sets versus the other two rate of spread methods.

>>>>

%=================================================================
% Full title of the paper (Capitalized)
\Title{A Multi-Fidelity Framework for Wildland Fire Behavior Simulations over Complex Terrain}

@@

There are various ways of simulating fire in vegetation which can depend on the fire behavior of interest. For example, for a fire spreading through surface vegetation (e.g., grass) the fire behavior differs depending on the orientation of the local wind relative to the direction of fire spread. For head fires, the wind flow and fire spread are in the same direction. Back fires spread into the wind. Flank fires spread approximately orthogonal to wind. For a surface fire spreading from a point in a uniform vegetation, the head fire portion of the resulting fireline has the largest flame base (easiest to resolve) and is characterized by fastest spread and flames with larger lengths that tilt toward the unburned vegetation. Backfires have the slowest spread and smallest flame base (require finer grid cells than the head fire) and have flames with smaller lengths that tilt away from the unburned vegetation. The roles of radiant and convective heat transfer differ along the fireline. Heat gain b
\begin{enumerate}
\item {\bf Particle Model:} The vegetation (surface and raised) is represented by a collection of Lagrangian particles that are heated via convection and radiation. This model, with sufficient grid resolution, is appropriate for head, back, and flank surface fires, as well as fire through raised vegetation (e.g., trees). Heat transfer in the volume containing the vegetation is modeled in all three directions. This model is appropriate for grid resolutions of the order of 1~m or less, depending on the size of the flame base and properties of the vegetation (e.g.,~\cite{Perez-Ramirez:FT2017}).
\item {\bf Boundary Fuel Model:} Surface vegetation has its own grid and is modeled like a porous solid with a thickness equal to the height of the vegetation. This model was designed for head fire spread in surface vegetation based on the assumption that heat transfer in the fuel bed is dominated by radiation from the overhead flame and therefore in the vertical direction.  In the implementation here, the height of the surface vegetation is assumed to unresolved on the grid. The appropriate gas-phase grid resolution of the order of 1~m to 10~m.
\item {\bf Level Set Method:} The fire front of a surface fire propagates using purely empirical rules in a level set method. Thermal degradation of the surface vegetation is not modeled. More than one implementation of this method is possible, largely differentiated by how the wind and fire-atmosphere interaction is modeled. The simpler implementations of this model can use grid resolutions coarser, 10 m and greater, than the more physics-based particle and boundary fuel models. The level set model can be used for firespread in the surface vegetation along with the particle model for fire behavior in raised vegetation. 
\end{enumerate}
The Particle Method and Boundary Fuel Model require thermo-physical properties of the vegetative fuels and the fire spread rate is {\em predicted} by the model. The Level Set Method relies on a set of predetermined spread rates for different types of vegetation and wind speeds. The pyrolysis model, developed specifically for vegetation~\cite{Porterie:2006,Morvan:CF2004,Houssami:2016}, consists of three reactions:

@@

In many simulations of wildland fire, the surface vegetation layer is too shallow to be resolved explicitly, as is done when using Lagrangian particles to represent the vegetation. In such cases, the surface vegetation can be modeled as a porous boundary consisting of a layer of dry vegetation, moisture, and air, underneath which is hard ground~\cite{Mell:IJWF2007}. The drag exerted by the vegetation is modeled using a special velocity boundary condition, and convective heat transfer is modeled via a source term in the one-dimensional heat conduction equation that is solved through the layer of vegetation and solid ground. Thermal radiation penetrates the vegetation layer via a 1-D radiative transport equation that is used for semi-transparent solids.

@@

For simulations of wildland fires spanning large areas that cannot be gridded finely enough to predict fire spread using the particle or boundary fuel models for vegetation, fire spread via a level approach is used~\cite{Bova:IJWF2015}. It's assumed that a surface fire spreading from a point under certain wind, slope, and vegetation conditions does so with an ellipse-shaped fire front with, for an effective wind vector, a fixed length-to-breadth ratio~\cite{Finney:FARSITE,Bova:IJWF2015}. Determining the spread rate vector at a point on the simulated fire front requires a formula for the magnitude of head fire rate of spread which is adjusted based on the location of the point on the idealized ellipse (e.g., head versus flank fire). This location is determined from the orientation of the effective wind vector relative to the normal to the fire front represented by the level set.

The location of a fire front at time $t$ is defined by the set of points $x,y$ where the level set scalar function $\phi(x,y,t)=0$. The advection equation for the level set function is
\begin{equation}
   \frac{D \phi(x,y,t)}{D t} =  \frac{\partial \phi}{\partial t} + \mathbf{R}_{uv} \cdot \nabla (\phi) = 0 \label{eqn:lset}
\end{equation}
where $\mathbf{R}_{uv}=(R_u,R_v)$ is the spread rate vector.

As stated above, the magnitude of the spread rate vector on a point on the level set fire front is determined by adjusting the magnitude of the head fire spread rate. The direction of the spread rate vector is assumed to be normal to the fire front. A formula for magnitude of the head fire spread rate can be obtained a number of ways, including empirical models~\cite{Cheney:IJWF1998} and physics-based simulations~\cite{Mell:FBFC2019}. A commonly used semi-empirical model is the Rothermel model~\cite{Rothermel:1972,Albini:1976} which can be implemented in FDS. Here, the magnitude of the head fire spread rate is
\begin{equation}
  \|\mathbf{R}_{uv}\|=R_0 \left(1 + \sqrt{(\boldsymbol{\phi}_W+\boldsymbol{\phi}_S) \cdot (\boldsymbol{\phi}_W+\boldsymbol{\phi}_S) } \right)
\end{equation}
where $R_0$ is the zero velocity, zero slope rate of spread that depends on the fuel properties. The vectors $\boldsymbol{\phi}_W,\boldsymbol{\phi}_S \ge 0$ are computed using the local wind and slope through empirical rules~\cite{Wilson:1980}. See Appendix 1 of Ref~\cite{Bova:IJWF2015} for details on these calculations. The Rothermel model has been used in other CFD-based approaches including the atmospheric weighted models WRF-SFIRE~\cite{Mandel:2014} and CAWFE~\cite{Coen:2}.

In FDS, the level set spread model can be implemented in the following ways:
\begin{enumerate}
\item Only the level set simulation is performed, with a constant and uniform specified wind and slope. The wind is not affected by the terrain, and there is no fire.
\item The wind field is established over the terrain, but it is ``frozen'' when the fire ignites.
\item The wind field follows the terrain, but there is no actual fire in the simulation, just front-tracking.  The level set evolves continuously in time with the flow field.
\item The wind and fire are fully-coupled and the resulting wind values are used in the head fire spread rate formula. When the fire front arrives at a given surface cell, it burns for a finite duration and with a heat release per unit area provided as part of the fuel model.
\item The wind and fire are fully-coupled in the gas-phased, but the head fire spread rate is not influenced by the wind speed. 
\end{enumerate}
When fully coupled to the CFD model, the level set function acts as an igniter as it reaches a new surface cell. The cell burns for a pre-determined amount of time. For cases where the depth of the fireline is less than the size of a grid cell, the burning time is extended while the burning rate is decreased so as to maintain conservation of fuel mass.

@@

For outdoor flows FDS prescribes the wind field through special ``open'' boundary conditions on the exterior of the computational domain.  The boundary conditions consist of three components:
\begin{enumerate}
\item[(i)] the specified upstream wind field (vertical profile of streamwise velocity components) based on prescribed Monin-Obukhov parameters imposed on fluid elements \emph{entering} the domain,
\item[(ii)] optional specification of upstream turbulence based on Jarrin's synthetic eddy method \cite{Jarrin:2008} (which is possible with the code, but not utilized in the test cases within this paper), and
\item[(iii)] nonuniform and nonstationary Dirichlet pressure boundary values for the Poisson equation.
\end{enumerate}
The following sections provide more detail on the implementation of the boundary profiles and the pressure boundary conditions.

@@

% \subsection{Turbulent Inflow Conditions}

% It is possible in FDS to superimpose synthetic turbulence on a specified inlet profile utilizing Jarrin's synthetic eddy method (SEM) \cite{Jarrin:2008}.  An example verification case is presented in Fig.~\ref{fig:semprofs}.  The left image of the figure shows contours of the normal component of velocity at the inlet.  The plot on the right shows the mean and rms profiles compared to the specified inlet parameters.  The domain is 20 m in the streamwise dimension by 12 m in the spanwise by 13 m in height.  In The turbulence intensity is set to 10 \% and the turbulence integral length scale is set to 3 m.  The specified MO parameters are $L=-667\;\mathrm{m}$, $z_0=0.022\;\mathrm{m}$, and $u_r=10\;\mathrm{m/s}$ and $T_r=20\;^\circ\mathrm{C}$ at reference height of 10 m.  The simulation runs for 60 seconds with statistics collected over the last 30 seconds.  The lateral, top, and downstream boundaries are set to ``open'' pressure boundaries as discussed in the next section.

% \begin{figure}[ht]
% \centering
% \includegraphics[width=0.5\textwidth,valign=c]{figures/sem_open_wind_contours.pdf}
% \includegraphics[width=0.48\textwidth,valign=c]{figures/sem_open_wind_u_prof.pdf}
% \caption{Example of inflow synthetic turbulence superimposed on a Monin-Obukhov boundary layer profile. Contours of the normal component of velocity at the domain inlet taken at a time midway through the simulation (left); mean and rms profiles along the inlet compared with specified profiles (right).}
% \label{fig:semprofs}
% \end{figure}

% Other approaches to imposing boundary layer turbulence are possible (e.g., \cite{Munoz-Esparza:2014}).  The cases presented later in this paper do not utilize SEM since detailed data on the turbulence statistics were not available.

@@

\begin{table}[ht]
\begin{center}
\caption[Measured properties for the CSIRO Grassland Fire cases]{Measured properties for the CSIRO Grassland Fire cases~\cite{Cheney:IJWF1993}.}
\label{Properties_Grasses}
\begin{tabular}{|l|c|c|c|}
\hline
Property                        & Units        & Case C064 & Case F19    \\ \hline \hline
Wind Speed                      & m/s          & 4.6       & 4.8         \\ \hline
Ambient Temperature             & $^\circ$C    & 32.       & 34          \\ \hline
Surface Area to Volume Ratio    & m$^{-1}$     & 9770      & 12240       \\ \hline
Grass Height                    & m            & 0.21      & 0.51        \\ \hline
Bulk Mass per Unit Area         & kg/m$^2$     & 0.283     & 0.313       \\ \hline
Moisture Fraction               & \%           & 6.3       & 5.8         \\ \hline
Measured RoS                    & m/s          & 1.0       & 1.5         \\ \hline
Calc'd RoS, Particle Method (0.25 m, 0.5 m, 1.0 m resolution) & m/s           & 1.4, 1.3, 1.4           & 1.1, 1.2, 1.2          \\ \hline
Calc'd RoS, Boundary Fuel Method (0.25 m, 0.5 m, 1.0 m resolution) & m/s      & 1.5, 1.4, 1.7           & 1.3, 1.3, 1.3           \\ \hline
Calc'd RoS, Level Set Method (5 m, 10 m, 20 m resolution) & m/s          & 1.1, 1.1, 1.2          & 0.4, 0.5, 0.8          \\ \hline
\end{tabular}
\end{center}
\end{table}

@@

The simulations have been conducted with the three different fire spread models, and each case has been run with three levels of spatial resolution. The simulations using the Particle and Boundary Fuel Models have been run with cubic grid cells that are 0.25~m, 0.5~m, and 1~m on a side. The simulations using the Level Set Model are run with cells of 5~m, 10~m, and 20~m. The predicted rates of spread for the two cases are listed in Table~\ref{Properties_Grasses} and shown graphically in Fig.~\ref{CSIRO}. For these simulations, the rate of spread predicted by the Particle and Boundary Fuel Models are comparable to the measured rate of spread while the level set method under-predicts the RoS. The under-prediction by the level set simulations is a consequence of the the limitations of the Rothermel model which is based on empirical relations largely derived from laboratory-scale experiments. It is well known that practical application of the Rothermel model often requires calibration for the case of interest~\cite{Arca_2007}. The use of the Rothermel for the head fire spread rate in surface vegetation in FDS is to be viewed a placeholder that is in-line with other CFD based models. If the observed head fire rate of spread for the F19 and C064 experiments is used in FDS, then good agreement with the observed fire perimeter is obtained (not shown). Further work is needed to develop rate of spread models appropriate for use with the Level Set Model (e.g.,~\cite{Mell:FBFC2019}).
